---
output: github_document

---
<style>
body {
text-align: justify}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r, echo = FALSE, out.width="25%", message = FALSE, warning = FALSE}

logo <- system.file("extdata", "GeRnika.png", package = "GeRnika")
knitr::include_graphics(logo)
```


`GeRnika` is an open-source R package that is capable of simulating, visualizing and comparing tumor evolution data by using simple commands. This aims at providing a tool to help researchers to easily simulate tumor clonal data and analyze the results of their approaches for studying the composition and the evolutionary history of tumors.

## Installation

`GeRnika` may be easily installed with the execution of the following command:

```{r, eval = FALSE}
# commands for installing the package from github
devtools::install_github("Aitorzan3/GeRnika")
```

Note that in order to install the vignettes together with the package, it is necessary to attach the packages "markdown" and "knitcitations" to your namespace. Once you have done that, you may use the following instruction to install de vignettes of `GeRnika`:

```{r, eval = FALSE}
# commands for installing the package together with its vignettes from github
devtools::install_github("Aitorzan3/GeRnika", build_vignettes = TRUE)
```

Once the package has been installed, the namespace of `GeRnika` may be loaded by using

```{r, eval=FALSE}
library(GeRnika)
```

This produces a short report about the versions of the packages that are used by `GeRnika`, providing information about any conflict related to its dependencies.

## Documentation

To view documentation for the version of this package installed in your system, start R and enter:

```{r, eval=FALSE}
browseVignettes("GeRnika")
```


## Design principles
Regarding the principles related to the design of `GeRnika`, this has been implemented in order to be fundamentally **intuitive** and **easy to use**. This is achieved by offering **accesible** methods for simulating and analyzing tumor phylogenies by using **simple commands**, **all in one** single package.


## Acknowledgments

`GeRnika` would not have been possible without the hard work of the *Intelligent Systems Group* members, specially **Borja Calvo** and **Maitena Tellaetxe**, who implemented various methods gathered in this package and coordinated its creation. 



